Fast Sorting on Modern Computers

### 2005-08-14 docs/tech_report/report.tex
some spell checks, removed unneccessary files, rename elementary
more minor edits.
minor changes mergeed in david's heap changes
changes for testing
fixed abstract and title page

\documentclass[dvips]{report}
\usepackage{ifthen}
\usepackage{varioref}
\usepackage{listings}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{float}
\usepackage{lgrind}

@@

\newcommand{\cc}[1]{\texttt{#1}}
\newcommand{\n}[1]{\textit{#1}}
\newcommand{\code}[2]{\begin{figure}\lstinputlisting{code/#2}\caption{#1}\label{#1}\end{figure}}

@@

\title{Fast Sorting on Modern Computers}
\author{Paul Biggar \and David Gregg}
\date{}

@@

\begin{titlepage}
        \begin{center}
                \vspace*{95mm}
                \LARGE Fast Sorting on Modern Computers\\
                \vspace{1em}
                \large Paul Biggar, David Gregg\\
                \large Department of Computer Science\\
				\large University of Dublin, Trinity College
        \end{center}
\end{titlepage}

\begin{abstract}
This report describes the creation and analysis of sorts designed to take
advantage of features of modern processors. The work on cache-conscious sorting
by LaMarca and Ladner is the starting point; this is implemented and analysed
using a hardware simulator. Results relating to a variety of cache and branch
predictor configurations are obtained and analysed for a variety of sorts,
beginning with insertion and selection sorts, bubblesort and shakersort, and
continuing to heapsort, mergesort, quicksort and radixsort. Actual and expected
results are compared together and to the results of LaMarca's research.
\end{abstract}

@@

\input ordernsquared.tex

>>>>

\documentclass{report}
\usepackage{ifthen}
\usepackage{varioref}
\usepackage{listings}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{float}
\usepackage{lgrind}

@@

%cc is used for
% - unix commans or programs: gcc, sim-bpred etc
\newcommand{\cc}[1]{\texttt{#1}}

%n is used for
% - proper names: Atom, SimpleScalar, Papi
\newcommand{\n}[1]{\textit{#1}}

\newcommand{\code}[2]{\begin{figure}\lstinputlisting{code/#2}\caption{#1}\label{#1}\end{figure}}

@@

@@

\begin{titlepage}
       \begin{center}
				\vspace*{40mm}
                \Huge Sorting in the Presence of Branch Prediction and Caches\\
                \vspace{1em}
                \Large Fast Sorting on Modern Computers\\
                \vspace{2em}
                \Large Paul Biggar \hspace{1cm} David Gregg\\
				\vspace{2em}
				\large Technical Report TR-CS-2005-XX\\
                \large Department of Computer Science,\\
				\large University of Dublin, Trinity College,\\
				\large Dublin 2, Ireland.
        \end{center}
\end{titlepage}

\begin{abstract}
Sorting is one of the most important and studied problems in computer
science. Many good algorithms exist which offer various trade-offs in
efficiency, simplicity and memory use. However most of these
algorithms were discovered decades ago at a time when computer
architectures were much simpler than today. Branch prediction and
cache memories are two developments in computer architecture that have
a particularly large impact on the performance of sorting algorithms.

This report describes a study of the behaviour of sorting algorithms
on branch predictors and caches. Our work on branch prediction is
almost entirely new, and finds a number of important results. In
particular we show that insertion sort causes the fewest branch
mispredictions of any comparison-based algorithm, that optimizations,
such as the choice of the pivot in quicksort, can have a large impact
on the predictability of branches, and that advanced two-level branch
predictors are usually worse at predicting branches in sorting
algorithms than simpler branch predictors. In many cases it is
possible to draw links between classical theoretical analyses of
algorithms and their branch prediction behaviour.

The other main work described in this report is an analysis of the
behaviour of sorting algorithms on modern caches. Over the last decade
there has been considerable interest in optimizing sorting algorithms
to reduce the number of cache misses. We experimentally study the
cache performance of both classical sorting algorithms, and a variety
of cache-optimized algorithms, primarily those proposed by LaMarca and
Ladner. Our experiments cover a much wider range of algorithms than
other work, including the $O(N^2)$ sorts and shellsort, all within a
single framework. We discover a number of new results, particularly
relating to the branch prediction behaviour of cache-optimized sorts.

We also developed a number of other improvements to the algorithms,
such as removing the need for a sentinel in heapsort. Overall, we
found that a cache-optimized radix sort was the fastest algorithm for
all large arrays; the absence of comparison branches means that the
algorithm causes almost no branch mispredictions.
\end{abstract}

@@

\input elementary.tex
